---
title: "NCAA2020M_EDA"
author: "Cameron Darling"
output: html_document
---

# Libraries

```{r message = FALSE, warning = FALSE}
rm(list = ls())
library(dplyr)
library(readr)
library(data.table)
```

# Import Data

```{r message = FALSE, warning = FALSE}
events15 <- read_csv('Data/MEvents2015.csv')
events16 <- read_csv('Data/MEvents2016.csv')
events17 <- read_csv('Data/MEvents2017.csv')
events18 <- read_csv('Data/MEvents2018.csv')
events19 <- read_csv('Data/MEvents2019.csv')
teams <- read_csv('Data/MTeams.csv')
```

```{r}
DayNum_Regular <- 133
```


```{r}
glimpse(events19)
```

```{r}
summary(events19)
```

# View All Possible Event Types in dataset

```{r}
unique(events15$EventType)
unique(events16$EventType)
unique(events17$EventType)
unique(events18$EventType)
unique(events19$EventType)
```

# View All Possible combination of EventType + EventSubType

```{r}
unique((events19 %>% mutate(check = paste(EventType,EventSubType)))$check)
```

# Create Metrics for Events Data using standard formula

```{r}
Standard_Metrics <- function(df) {
  df %>% 
    mutate(TeamID = WTeamID,
           Opp_TeamID = LTeamID,
           FGM_2P = ifelse(WTeamID == EventTeamID,
                           ifelse(EventType == 'made2',1,0),0),
           FGA_2P = ifelse(WTeamID == EventTeamID,
                         ifelse(EventType == 'made2',1,
                         ifelse(EventType == 'miss2',1,0)),0),
           FGM_3P = ifelse(WTeamID == EventTeamID,
                         ifelse(EventType == 'made3',1,0),0),
           FGA_3P = ifelse(WTeamID == EventTeamID,
                         ifelse(EventType == 'made3',1,
                         ifelse(EventType == 'miss3',1,0)),0),
           FTM = ifelse(WTeamID == EventTeamID,
                      ifelse(EventType == 'made1',1,0),0),
           FTA = ifelse(WTeamID == EventTeamID,
                      ifelse(EventType == 'made1',1,
                      ifelse(EventType == 'miss1',1,0)),0),
           AST = ifelse(WTeamID == EventTeamID,
                      ifelse(EventType == 'assist',1,0),0),
           TO = ifelse(WTeamID == EventTeamID,
                     ifelse(EventType == 'turnover',1,0),0),
           ORB = ifelse(WTeamID == EventTeamID,
                      ifelse(EventType == 'reb', 
                      ifelse(EventSubType == 'off',1,0),0),0),
           DRB = ifelse(WTeamID == EventTeamID,
                      ifelse(EventType == 'reb',
                      ifelse(EventSubType == 'def',1,0),0),0),
           STL = ifelse(WTeamID == EventTeamID,
                      ifelse(EventType == 'steal',1,0),0),
           BLK = ifelse(WTeamID == EventTeamID,
                      ifelse(EventType == 'block',1,0),0),
           FOUL = ifelse(WTeamID == EventTeamID,
                       ifelse(EventType == 'foul',1,0),0),
           Pts = ifelse(WTeamID == EventTeamID, 2 * FGM_2P + 3 * FGM_3P + FTM, 0),
           Opp_FGM_2P = ifelse(LTeamID == EventTeamID,
                           ifelse(EventType == 'made2',1,0),0),
           Opp_FGA_2P = ifelse(LTeamID == EventTeamID,
                         ifelse(EventType == 'made2',1,
                         ifelse(EventType == 'miss2',1,0)),0),
           Opp_FGM_3P = ifelse(LTeamID == EventTeamID,
                         ifelse(EventType == 'made3',1,0),0),
           Opp_FGA_3P = ifelse(LTeamID == EventTeamID,
                         ifelse(EventType == 'made3',1,
                         ifelse(EventType == 'miss3',1,0)),0),
           Opp_FTM = ifelse(LTeamID == EventTeamID,
                      ifelse(EventType == 'made1',1,0),0),
           Opp_FTA = ifelse(LTeamID == EventTeamID,
                      ifelse(EventType == 'made1',1,
                      ifelse(EventType == 'miss1',1,0)),0),
           Opp_AST = ifelse(LTeamID == EventTeamID,
                      ifelse(EventType == 'assist',1,0),0),
           Opp_TO = ifelse(LTeamID == EventTeamID,
                     ifelse(EventType == 'turnover',1,0),0),
           Opp_ORB = ifelse(LTeamID == EventTeamID,
                      ifelse(EventType == 'reb', 
                      ifelse(EventSubType == 'off',1,0),0),0),
           Opp_DRB = ifelse(LTeamID == EventTeamID,
                      ifelse(EventType == 'reb',
                      ifelse(EventSubType == 'def',1,0),0),0),
           Opp_STL = ifelse(LTeamID == EventTeamID,
                      ifelse(EventType == 'steal',1,0),0),
           Opp_BLK = ifelse(LTeamID == EventTeamID,
                      ifelse(EventType == 'block',1,0),0),
           Opp_FOUL = ifelse(LTeamID == EventTeamID,
                       ifelse(EventType == 'foul',1,0),0),
           Opp_Pts = ifelse(LTeamID == EventTeamID, 2 * Opp_FGM_2P + 3 * Opp_FGM_3P + Opp_FTM, 0))
}
```

# Metric for determining game part

```{r}
GamePart <- function(df) {
  df %>% 
    mutate(Game_Part = ifelse(ElapsedSeconds <= 60 * 20, '1H',
                   ifelse(ElapsedSeconds <= 60 * 40, '2H',
                          'OT')))
}
```

# Identify the Games that do not match between event points and total points


# Break down data into 1H/2H/Total Per Team

```{r}
BreakDown_GamePart <- function(df) {
  first_half <- Standard_Metrics(df) %>% 
           GamePart() %>% 
           filter(Game_Part == '1H') %>%
           group_by(TeamID,Opp_TeamID,DayNum) %>% 
           summarise(FinalScore = max(WFinalScore),
                     Opp_FinalScore = max(LFinalScore),
                     FGM_2P = sum(FGM_2P),
                     FGA_2P = sum(FGA_2P),
                     FGM_3P = sum(FGM_3P),
                     FGA_3P = sum(FGA_3P),
                     FTM = sum(FTM),
                     FTA = sum(FTA),
                     AST = sum(AST),
                     TO = sum(TO),
                     ORB = sum(ORB),
                     DRB = sum(DRB),
                     STL = sum(STL),
                     BLK = sum(BLK),
                     FOUL = sum(FOUL),
                     Pts = sum(Pts),
                     Opp_FGM_2P = sum(Opp_FGM_2P),
                     Opp_FGA_2P = sum(Opp_FGA_2P),
                     Opp_FGM_3P = sum(Opp_FGM_3P),
                     Opp_FGA_3P = sum(Opp_FGA_3P),
                     Opp_FTM = sum(Opp_FTM),
                     Opp_FTA = sum(Opp_FTA),
                     Opp_AST = sum(Opp_AST),
                     Opp_TO = sum(Opp_TO),
                     Opp_ORB = sum(Opp_ORB),
                     Opp_DRB = sum(Opp_DRB),
                     Opp_STL = sum(Opp_STL),
                     Opp_BLK = sum(Opp_BLK),
                     Opp_FOUL = sum(Opp_FOUL),
                     Opp_Pts = sum(Opp_Pts)) %>% 
           ungroup() %>% 
           filter(Pts != 0) %>% 
           mutate(POSS = .96 * ((FGA_2P + FGA_3P) + TO + (.475 * FTA) - ORB),
                  Opp_POSS = .96 * ((Opp_FGA_2P + Opp_FGA_3P) + TO + (.475 * Opp_FTA) - Opp_ORB)) %>% 
           mutate(Win = ifelse(FinalScore > Opp_FinalScore,1,0),
                  Loss = ifelse(FinalScore < Opp_FinalScore,1,0),
                  Tie = ifelse(FinalScore == Opp_FinalScore,1,0))
  first_half_flipped <- first_half
  flipped_names <- c('Opp_TeamID', 'TeamID', 'DayNum','Opp_FinalScore', 'FinalScore','Opp_FGM_2P','Opp_FGA_2P','Opp_FGM_3P','Opp_FGA_3P',
                     'Opp_FTM','Opp_FTA','Opp_AST','Opp_TO','Opp_ORB','Opp_DRB','Opp_STL','Opp_BLK','Opp_FOUL','Opp_Pts',
                     'FGM_2P','FGA_2P','FGM_3P','FGA_3P','FTM','FTA','AST','TO','ORB','DRB','STL','BLK','FOUL','Pts','Opp_POSS','POSS','Loss','Win','Tie')
  names(first_half_flipped) <- flipped_names
  first_half_flipped <- first_half_flipped %>% 
    select(TeamID,Opp_TeamID,DayNum,Win,Loss,Tie,FinalScore,Opp_FinalScore,FGM_2P,FGA_2P,FGM_3P,FGA_3P,FTM,FTA,AST,TO,ORB,DRB,STL,BLK,FOUL,Pts,Opp_FGM_2P,
           Opp_FGA_2P,Opp_FGM_3P,Opp_FGA_3P,Opp_FTM,Opp_FTA,Opp_AST,Opp_TO,Opp_ORB,Opp_DRB,Opp_STL,Opp_BLK,Opp_FOUL,Opp_Pts,POSS,Opp_POSS)
  first_half_combined <- rbind(first_half,first_half_flipped) %>% 
           arrange(TeamID,DayNum) %>% 
           group_by(TeamID) %>% 
           mutate(Game = seq(n())) %>% 
    ungroup()
  second_half <- Standard_Metrics(df) %>% 
           GamePart() %>% 
           filter(Game_Part == '2H') %>% 
           group_by(TeamID,Opp_TeamID,DayNum) %>% 
           summarise(FinalScore = max(WFinalScore),
                     Opp_FinalScore = max(LFinalScore),
                     FGM_2P = sum(FGM_2P),
                     FGA_2P = sum(FGA_2P),
                     FGM_3P = sum(FGM_3P),
                     FGA_3P = sum(FGA_3P),
                     FTM = sum(FTM),
                     FTA = sum(FTA),
                     AST = sum(AST),
                     TO = sum(TO),
                     ORB = sum(ORB),
                     DRB = sum(DRB),
                     STL = sum(STL),
                     BLK = sum(BLK),
                     FOUL = sum(FOUL),
                     Pts = sum(Pts),
                     Opp_FGM_2P = sum(Opp_FGM_2P),
                     Opp_FGA_2P = sum(Opp_FGA_2P),
                     Opp_FGM_3P = sum(Opp_FGM_3P),
                     Opp_FGA_3P = sum(Opp_FGA_3P),
                     Opp_FTM = sum(Opp_FTM),
                     Opp_FTA = sum(Opp_FTA),
                     Opp_AST = sum(Opp_AST),
                     Opp_TO = sum(Opp_TO),
                     Opp_ORB = sum(Opp_ORB),
                     Opp_DRB = sum(Opp_DRB),
                     Opp_STL = sum(Opp_STL),
                     Opp_BLK = sum(Opp_BLK),
                     Opp_FOUL = sum(Opp_FOUL),
                     Opp_Pts = sum(Opp_Pts)) %>%  
           ungroup() %>% 
           filter(Pts != 0) %>% 
           mutate(POSS = .96 * ((FGA_2P + FGA_3P) + TO + (.475 * FTA) - ORB),
                  Opp_POSS = .96 * ((Opp_FGA_2P + Opp_FGA_3P) + TO + (.475 * Opp_FTA) - Opp_ORB)) %>% 
           mutate(Win = ifelse(FinalScore > Opp_FinalScore,1,0),
                  Loss = ifelse(FinalScore < Opp_FinalScore,1,0),
                  Tie = ifelse(FinalScore == Opp_FinalScore,1,0))
  second_half_flipped <- second_half
  names(second_half_flipped) <- flipped_names
  second_half_flipped <- second_half_flipped %>% 
    select(TeamID,Opp_TeamID,DayNum,Win,Loss,Tie,FinalScore,Opp_FinalScore,FGM_2P,FGA_2P,FGM_3P,FGA_3P,FTM,FTA,AST,TO,ORB,DRB,STL,BLK,FOUL,Pts,Opp_FGM_2P,
           Opp_FGA_2P,Opp_FGM_3P,Opp_FGA_3P,Opp_FTM,Opp_FTA,Opp_AST,Opp_TO,Opp_ORB,Opp_DRB,Opp_STL,Opp_BLK,Opp_FOUL,Opp_Pts,POSS,Opp_POSS)
  second_half_combined <- rbind(second_half,second_half_flipped) %>% 
           arrange(TeamID,DayNum) %>% 
           group_by(TeamID) %>% 
           mutate(Game = seq(n())) %>% 
    ungroup()
  full_game <- Standard_Metrics(df) %>% 
           GamePart() %>% 
           group_by(TeamID,Opp_TeamID,DayNum) %>% 
           summarise(FinalScore = max(WFinalScore),
                     Opp_FinalScore = max(LFinalScore),
                     FGM_2P = sum(FGM_2P),
                     FGA_2P = sum(FGA_2P),
                     FGM_3P = sum(FGM_3P),
                     FGA_3P = sum(FGA_3P),
                     FTM = sum(FTM),
                     FTA = sum(FTA),
                     AST = sum(AST),
                     TO = sum(TO),
                     ORB = sum(ORB),
                     DRB = sum(DRB),
                     STL = sum(STL),
                     BLK = sum(BLK),
                     FOUL = sum(FOUL),
                     Pts = sum(Pts),
                     Opp_FGM_2P = sum(Opp_FGM_2P),
                     Opp_FGA_2P = sum(Opp_FGA_2P),
                     Opp_FGM_3P = sum(Opp_FGM_3P),
                     Opp_FGA_3P = sum(Opp_FGA_3P),
                     Opp_FTM = sum(Opp_FTM),
                     Opp_FTA = sum(Opp_FTA),
                     Opp_AST = sum(Opp_AST),
                     Opp_TO = sum(Opp_TO),
                     Opp_ORB = sum(Opp_ORB),
                     Opp_DRB = sum(Opp_DRB),
                     Opp_STL = sum(Opp_STL),
                     Opp_BLK = sum(Opp_BLK),
                     Opp_FOUL = sum(Opp_FOUL),
                     Opp_Pts = sum(Opp_Pts)) %>%  
           ungroup() %>% 
           filter(Pts != 0) %>% 
           mutate(POSS = .96 * ((FGA_2P + FGA_3P) + TO + (.475 * FTA) - ORB),
                  Opp_POSS = .96 * ((Opp_FGA_2P + Opp_FGA_3P) + TO + (.475 * Opp_FTA) - Opp_ORB)) %>% 
           mutate(Win = ifelse(FinalScore > Opp_FinalScore,1,0),
                  Loss = ifelse(FinalScore < Opp_FinalScore,1,0),
                  Tie = ifelse(FinalScore == Opp_FinalScore,1,0))
  full_game_flipped <- full_game
  names(full_game_flipped) <- flipped_names
  full_game_flipped <- full_game_flipped %>% 
    select(TeamID,Opp_TeamID,DayNum,Win,Loss,Tie,FinalScore,Opp_FinalScore,FGM_2P,FGA_2P,FGM_3P,FGA_3P,FTM,FTA,AST,TO,ORB,DRB,STL,BLK,FOUL,Pts,Opp_FGM_2P,
           Opp_FGA_2P,Opp_FGM_3P,Opp_FGA_3P,Opp_FTM,Opp_FTA,Opp_AST,Opp_TO,Opp_ORB,Opp_DRB,Opp_STL,Opp_BLK,Opp_FOUL,Opp_Pts,POSS,Opp_POSS)
  full_game_combined <- rbind(full_game,full_game_flipped) %>% 
           arrange(TeamID,DayNum) %>% 
           group_by(TeamID) %>% 
           mutate(Game = seq(n())) %>% 
    ungroup()
  assign(paste(deparse(substitute(df)),'_1H', sep = ''),
         first_half_combined,
         envir = .GlobalEnv)
  assign(paste(deparse(substitute(df)),'_2H', sep = ''),
         second_half_combined, 
         envir = .GlobalEnv)
  assign(paste(deparse(substitute(df)),'_TOT', sep = ''),
         full_game_combined, 
         envir = .GlobalEnv)
}
```

```{r}
BreakDown_GamePart(events15)
BreakDown_GamePart(events16)
BreakDown_GamePart(events17)
BreakDown_GamePart(events18)
BreakDown_GamePart(events19)
```


# Check Games that Scores do not line up

```{r}
error_check <- function(df) {
  df %>% filter(FinalScore != Pts | Opp_FinalScore != Opp_Pts) %>% select(TeamID, Opp_TeamID, DayNum, FinalScore, Opp_FinalScore, Pts, Opp_Pts)
}
```

```{r}
error_check(events15_TOT)
```

```{r}
error_game <- events15 %>% filter(WTeamID == 1164 & LTeamID == 1288 & DayNum == 117) %>% Standard_Metrics()
```

# Calculate CPI and Offense/Defensive Factor

```{r}
cpi_calculation <- function(df) {
  step1 <- df %>% select(TeamID, Win, Loss, Tie, FinalScore, Opp_FinalScore, POSS, Opp_POSS) %>% 
    group_by(TeamID) %>% 
    summarise(opp_opp_wins = sum(Win, na.rm = T), opp_opp_games = n(),
              opp_opp_ps = sum(FinalScore, na.rm = T), opp_opp_pa = sum(Opp_FinalScore, na.rm = T),
              opp_opp_poss = sum(POSS, na.rm = T),
              opp_opp_def_poss = sum(Opp_POSS, na.rm = T)) %>% 
    ungroup()
  step2 <- merge(df, step1, all.x = T, by.x = c('Opp_TeamID'), by.y = c('TeamID')) %>% 
    select(TeamID, Opp_TeamID, Win, FinalScore, Opp_FinalScore, opp_opp_wins, opp_opp_games, opp_opp_ps, opp_opp_pa, opp_opp_poss, opp_opp_def_poss, POSS, Opp_POSS) %>% 
    group_by(TeamID) %>% 
    summarise(opp_wins = sum(Win), opp_games = n(), opp_opp_wins = sum(opp_opp_wins, na.rm = T), opp_opp_games = sum(opp_opp_games, na.rm = T),
              opp_ps = sum(FinalScore, na.rm = T), opp_pa = sum(Opp_FinalScore, na.rm = T),
              opp_opp_ps = sum(opp_opp_ps, na.rm = T), opp_opp_pa = sum(opp_opp_pa, na.rm = T),
              opp_opp_poss = sum(opp_opp_poss, na.rm = T), opp_poss = sum(POSS, na.rm = T),
              opp_opp_def_poss = sum(opp_opp_def_poss, na.rm = T), opp_def_poss = sum(Opp_POSS, na.rm = T)) %>% 
    ungroup()
  cpi <- merge(df,step2, all.x = T, by.x = c('Opp_TeamID'), by.y = c('TeamID')) %>% 
    select(TeamID, Opp_TeamID, Win, opp_wins, opp_games, opp_opp_wins, opp_opp_games, opp_opp_ps, opp_ps, opp_opp_pa, opp_pa, opp_opp_poss, opp_poss, 
           opp_opp_def_poss, opp_def_poss, POSS, Opp_POSS) %>% 
    group_by(TeamID) %>% 
    summarise(win_pct = (sum(Win, na.rm = T) / n()), opp_win_pct = (sum(opp_wins, na.rm = T) / sum(opp_games, na.rm = T)),
              opp_opp_win_pct = (sum(opp_opp_wins, na.rm = T) / sum(opp_opp_games, na.rm = T)),
              opp_opp_ps_avg = (sum(opp_opp_ps, na.rm = T) / sum(opp_opp_games, na.rm = T)),
              opp_opp_pa_avg = (sum(opp_opp_pa, na.rm = T) / sum(opp_opp_games, na.rm = T)),
              opp_ps_avg = (sum(opp_ps, na.rm = T) / sum(opp_games, na.rm = T)),
              opp_pa_avg = (sum(opp_pa, na.rm = T) / sum(opp_games, na.rm = T)),
              opp_opp_poss_avg = 1) %>% 
    mutate(cpi_value = round(.25 * win_pct + .5 * opp_win_pct + .25 * opp_opp_win_pct,3), cpi_rank = rank(1 - cpi_value, ties.method = 'first'),
           def_factor = round((opp_pa_avg - opp_opp_ps_avg) / opp_opp_ps_avg,3),
           off_factor = round((opp_ps_avg - opp_opp_pa_avg) / opp_opp_pa_avg,3)) %>% 
    ungroup() %>% 
    select(TeamID, cpi_value, cpi_rank, def_factor, off_factor) %>% 
    arrange(cpi_rank)
  
  opp_cpi <- cpi %>% select(Opp_TeamID = TeamID, opp_cpi_value = cpi_value, opp_cpi_rank = cpi_rank,
                                opp_def_factor = def_factor, opp_off_factor = off_factor)
  return(cpi)
}
```

```{r}
cpi_test <- cpi_calculation(events15_TOT %>% filter(DayNum <= DayNum_Regular)) %>% merge(teams) %>%
  arrange(cpi_rank)
cpi_test
```

```{r}
daily_cpi <- function(df) {
  min_daynum <- min(df %>% filter(Game == 7) %>% select(DayNum))
  combined_daily <- na.omit(data.frame(matrix(rep(NA,1), ncol = length(names(cpi_calculation(df))))) %>% mutate(DayNum = NA))
  for (i in min_daynum:DayNum_Regular) {
    temp_cpi <- df %>% 
      filter(DayNum <= i) %>% 
      cpi_calculation() %>% 
      mutate(DayNum = i + 1) %>% 
      select(TeamID,DayNum,cpi_value, cpi_rank, def_factor, off_factor)
    combined_daily <- rbind(temp_cpi,combined_daily) %>% 
      arrange(TeamID, DayNum)
  }
  return(combined_daily)
}

```

```{r}
dailycpi_15 <- daily_cpi(events15_TOT)
dailycpi_16 <- daily_cpi(events16_TOT)
dailycpi_17 <- daily_cpi(events17_TOT)
dailycpi_18 <- daily_cpi(events18_TOT)
dailycpi_19 <- daily_cpi(events19_TOT)
```


```{r}
min(events15_TOT %>% filter(Game == 7) %>% select(DayNum))
```

```{r}
previous_game_average <- function(df,g) {
  max_games <- max(df$Game)
  temp_avg <- na.omit(data.frame(matrix(rep(NA,1), 
                                        ncol = length(names(df %>% 
                                                              filter(Game <= 1) %>% 
                                                              group_by(TeamID) %>% 
                                                              mutate(maxgame = max(Game)) %>% 
                                                              summarise_all(mean) %>% 
                                                              select(-Opp_TeamID,-DayNum,-Win,-Loss,-Tie)
                                                            )))))
  names(temp_avg) <- paste('Avg_',names(temp_avg),sep = '')
  colnames(temp_avg)[names(temp_avg) == 'Avg_TeamID'] <- 'TeamID'
  colnames(temp_avg)[names(temp_avg) == 'Avg_Game'] <- 'Game'
  for (i in  1:max_games) {
      t <- df %>%  
        group_by(TeamID) %>% 
        mutate(maxgame = max(Game)) %>% 
        filter(Game <= i & Game > (i - ifelse(g == 50,max(Game),g))) %>%
        summarise_all(funs(round(mean(.),2))) %>% 
        mutate(Game = i) %>% 
        filter(Game == i) %>% 
        select(-Opp_TeamID,-DayNum,-Win,-Loss,-Tie)
      names(t) <- paste('Avg_',names(t),sep = '')
      colnames(t)[names(t) == 'Avg_TeamID'] <- 'TeamID'
      colnames(t)[names(t) == 'Avg_Game'] <- 'Game'
      temp_avg <- rbind(temp_avg,t) %>% arrange(TeamID, Game) %>% 
        group_by(TeamID) %>% 
        filter(Game >= ifelse(g == 50,Avg_maxgame,g) & Game <= Avg_maxgame) %>%
        ungroup()
  }
  return(temp_avg %>% select(-Avg_maxgame))
}
```
 
 
```{r}
# Input is full game level details and number of previous Games. For Full Season use 50
s <- previous_game_average(events15_TOT,7)
glimpse(s)
```
 
