---
title: "NCAA2020M_EDA"
author: "Cameron Darling"
output: html_document
---

# Libraries

```{r message = FALSE, warning = FALSE}
rm(list = ls())
library(dplyr)
```

# Import Data

```{r}
events15 <- read.csv('Data/MEvents2015.csv')
events16 <- read.csv('Data/MEvents2016.csv')
events17 <- read.csv('Data/MEvents2017.csv')
events18 <- read.csv('Data/MEvents2018.csv')
events19 <- read.csv('Data/MEvents2019.csv')
```

```{r}
glimpse(events19)
```

```{r}
summary(events19)
```

# View All Possible Event Types in dataset

```{r}
unique(events15$EventType)
unique(events16$EventType)
unique(events17$EventType)
unique(events18$EventType)
unique(events19$EventType)
```

# View All Possible combination of EventType + EventSubType

```{r}
unique((events19 %>% mutate(check = paste(EventType,EventSubType)))$check)
```

# Create Metrics for Events Data using standard formula

```{r}
Standard_Metrics <- function(df) {
  df %>% 
    mutate(FGM_2P = ifelse(EventType == 'made2',1,0),
         FGA_2P = ifelse(EventType == 'made2',1,
                         ifelse(EventType == 'miss2',1,0)),
         FGM_3P = ifelse(EventType == 'made3',1,0),
         FGA_3P = ifelse(EventType == 'made3',1,
                         ifelse(EventType == 'miss3',1,0)),
         FTM = ifelse(EventType == 'made1',1,0),
         FTA = ifelse(EventType == 'made1',1,
                      ifelse(EventType == 'miss1',1,0)),
         AST = ifelse(EventType == 'assist',1,0),
         TO = ifelse(EventType == 'turnover',1,0),
         ORB = ifelse(EventType == 'reb', 
                      ifelse(EventSubType == 'off',1,0),0),
         DRB = ifelse(EventType == 'reb',
                      ifelse(EventSubType == 'def',1,0),0),
         STL = ifelse(EventType == 'steal',1,0),
         BLK = ifelse(EventType == 'block',1,0),
         FOUL = ifelse(EventType == 'foul',1,0),
         PTS = 2 * FGM_2P + 3 * FGM_3P + FTM,
         WPts = ifelse(WTeamID == EventTeamID, PTS, 0),
         LPts = ifelse(LTeamID == EventTeamID, PTS, 0))
}
```

# Metric for determining game part

```{r}
GamePart <- function(df) {
  df %>% 
    mutate(Game_Part = ifelse(ElapsedSeconds <= 60 * 20, '1H',
                   ifelse(ElapsedSeconds <= 60 * 40, '2H',
                          'OT')))
}
```


# Formula for Advanced Metrics based on Game/Half Level Stats

```{r}
Advanced_Metrics <- function(df) {
  df %>% mutate(POSS = .96 * ((FGA_2P + FGA_3P) + TO + (.475 * FTA) - ORB),
                ORtg = 100 * (PTS / POSS),
                DRtg = 100 * 1)
}
```


# Identify the Games that do not match between event points and total points

```{r}
event_check <- Standard_Metrics(events19) %>% 
  group_by(WTeamID, LTeamID, DayNum) %>% 
  summarise(WFinalScore = max(WFinalScore),
            LFinalScore = max(LFinalScore),
            WPts = sum(WPts),
            LPts = sum(LPts)) %>% 
  ungroup() %>% 
  mutate(check = ifelse(WFinalScore == WPts & LFinalScore == LPts,"Yes","No")) %>% 
  filter(check != 'Yes')
  
```

# Adding in Event Statistics

```{r}
events19_2 <- Standard_Metrics(events19) %>% GamePart()
```

